---
title: "STAT 451 - Visualizing Data - Autumn 2025"
author: "Ariane Ducellier"
date: "09/30/2025"
output: pdf_document
---

# Tutorial Tidyverse part 1

Today, we are going to review useful R functions for reading and exploring data sets. The objective is to make sure that you have all the tools you need to explore your data set before starting to visualize it. For more on data exploration with R, I recommend the two books:
  
Data Wrangling with R Load, Explore, Transform and Visualize Data for Modeling with Tidyverse Libraries.
Santos, Gustavo R.  
2023; Birmingham : Packt Publishing, Limited
  
R for data science : import, tidy, transform, visualize, and model data
Wickham, Hadley, author.  
2023; Sebastopol, CA : O'Reilly Media, Inc.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We will need the following R libraries:

```{r libraries, message=FALSE}
library(tidyverse)
library(skimr)
```

## 1. Read data (csv format)

Read csv files with basic R.

```{r read_basic}
df <- read.csv("../../data/gapminder-data.csv")
print(class(df))
```

Read csv files with tidyverse.

```{r read_csv}
df_t <- read_csv("../../data/gapminder-data.csv")
print(class(df_t))
```

Read select columns, specify the encoding of the missing values.

```{r read_csv_sub}
df_t_sub <- read_csv("../../data/gapminder-data.csv",
  col_select=c("Country", "Year", "gdp_per_capita"),
  na=c("", "NA"))
```

## 2. Get basic information on the data.

Reads the first part of a data frame.

```{r head}
df_t <- read_csv("../../data/gapminder-data.csv")
head(df_t, 3)
```

Reads the last part of a data frame.

```{r tail}
tail(df_t, 3)
```

Gets column specifications of a tibble.

```{r spec}
spec(df_t)
```

Prints the data: number of rows and columns, type of columns, and first rows.

```{r glimpse}
glimpse(df_t)
```

Returns descriptive statistics on each column of a data frame.

```{r summary}
summary(df_t)
```

Provides a broad overview of a data frame, handles data of all types, dispatching a different set of summary functions based on the types of columns in the data frame.

```{r skim}
# Comment this line for knitting with pdf
#skim(df_t)
```

## 3. The pipe operator

The magrittr package provides the %>% operator as a shortcut for modifying an object in place. It is installed by default when installing tidyverse.

```{r iris}
data(iris)
df_iris <- iris %>%
  group_by(Species) %>%
  summarize_if(is.numeric, mean) %>%
  ungroup() %>%
  gather(measure, value, -Species) %>%
  arrange(value)
```

is the same as:

```{r iris_alt}
data(iris)
df_iris_alt <- group_by(iris, Species)
df_iris_alt <- summarize_if(df_iris_alt, is.numeric, mean)
df_iris_alt <- ungroup(df_iris_alt)
df_iris_alt <- gather(df_iris_alt, measure, value, -Species)
df_iris_alt <- arrange(df_iris_alt, value)
```

## 4. Transform the data

We will review some basic transformations.

```{r get_data}
header <- c("age", "workclass", "fnlwgt", "education",
  "education_num", "marital_status", "occupation",
  "relationship", "race", "sex", "capital_gain",
  "capital_loss", "hours_per_week", "native_country", "target")
df <- read_csv("https://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.data",
  col_names=header, trim_ws=TRUE)
```

### 4.1 Slicing

Cuts unwanted parts of the data set.

```{r slice}
df %>% slice_head(n=5)
df %>% slice_tail(n=5)
df %>% slice_min(age, prop=0.10)
df %>% slice_max(age, prop=0.30)
df %>% slice_sample(n=10, replace=TRUE)
```

### 4.2 Filtering

Apply a condition to one of the variables to filter unwanted rows of the data.

```{r filter}
df %>% filter(age > 30)
```

### 4.3 Selecting

Select specific columns.

```{r select}
df %>% select(marital_status, age)
```

### 4.4 Unique values

Get unique entries for categorical variables.

```{r distinct}
df %>% distinct(sex)
```

### 4.5 Grouping

Group by column and summarize.

```{r group}
df %>%
  group_by(workclass) %>%
  summarize(age_avg=mean(age))
```

### 4.6 Summarizing

The summary may be:
- counting observations
- counting available observations (i.e. not NA)
- getting first or last value
- compute statistics on each group (mean, standard deviation, quantile)

```{r summarize}
df %>% group_by(workclass) %>% summarize(n())
df %>% summarize(sum(!is.na(workclass)))
df %>% group_by(workclass) %>% summarize(first(age))
df %>% group_by(workclass) %>% summarize(sd(capital_gain))
df %>% group_by(workclass) %>% summarize(quantile(age, 0.5))
```

We can also apply the summary over selected columns.

```{r summarize_columns}
df %>% select(1, 3, 5, 11, 12, 13) %>% summarize(across(everything(), mean))
```

### 4.7 Arranging

To sort the data set.

```{r arrange}
df %>% arrange(native_country)
```

It is most useful to sort the data set after grouping and summarizing.

```{r arrange_group}
df %>%
  group_by(education) %>%
  summarize(count=n(),
            avg_net_gain = mean(capital_gain - capital_loss)) %>%
  arrange(desc(avg_net_gain))
```

### 4.8 Separating and uniting

This is often useful to create new columns.

```{r separate}
df %>% separate_wider_regex(cols=target, patterns=c(sign="(?:<=|>)", amount="\\d+", unit="."))
```

```{r unite}
df %>% unite(sex, race, age, col="description", sep="_", remove=FALSE)
```

### 4.8 Mutate function

This function can be used for many purposes. You can create new variables:

```{r mutate_add_new}
df %>%
  mutate(total_gain = capital_gain - capital_loss,
                tax = ifelse(total_gain >= 15000,
                             total_gain * 0.1,
                             0)
  ) %>%
  arrange(desc(tax))
```

We can use the mutate function to replace values (e.g. "?" by NA). Let see how many "?" we have in this data set.

```{r look_question_marks}
for (variable in colnames(df)) {
  print (
    paste (variable, dim(drop_na(df[df[variable]=="?", variable]))[1])
  )
}
```

We now replace the "?" by NA in the three columns workclass, occupation and native_country.

```{r replace}
df_replaced <- df %>%
  mutate(workclass = replace(workclass, workclass=="?", NA),
         occupation = replace(occupation, occupation=="?", NA),
         native_country =replace(native_country, native_country=="?", NA)
  )
```

Let us check that all the "?" have been replaced.

```{r look_question_marks_new}
for (variable in colnames(df_replaced)) {
  print (
    paste (variable, dim(drop_na(df_replaced[df_replaced[variable]=="?", variable]))[1])
  )
}
```

An alternative is to use the na_if function.

```{r na_if}
df %>% mutate(workclass = na_if(workclass, "?"),
              occupation = na_if(occupation, "?"),
              native_country = na_if(native_country, "?"))
```

Here are additional examples of the use of the mutate function. We can create a new column based on the values of another column:

```{r mutate_ex1}
df %>%
  mutate(over_under = case_match(target,
                                "<=50K" ~ "under",
                                ">50K" ~ "over")) %>%
  select(target, over_under)
```

or cut the range of a variable into intervals:

```{r mutate_ex2}
df %>%
  mutate(age_avg = mean(age),
         over_under_age_avg = cut(age,
                                  c(0, mean(age), max(age)),
                                  c("Lower than avg", "Above the avg"))
  ) %>%
  select(age, age_avg, over_under_age_avg)
```

### 4.9 Joining tibbles

Joining two data sets is a very useful and common function. We are going to use these 3 small data sets as an example.

```{r data}
sales <- data.frame(
  date = c("2022-01-01", "2022-01-02", "2022-01-03", "2022-01-04", "2022-01-05"),
  store_cd= c(1, 2, 3, 4, 5),
  product_cd= c(1, 2, 3, 4, 5),
  qty= c(10, 12, 9, 12, 8), 
  sales= c(30, 60, 45, 24, 32)
)

stores <- data.frame(
  store_cd= c(1, 2, 3, 4, 6),
  address= c("1 main st", "20 side st", "19 square blvd", "101 first st", "1002 retail ave"),
  city= c("Main", "East", "West", "North", "South"), 
  open_hours= c("7-23", "7-23", "9-21", "9-21", "9-21")
  )

products <- data.frame(
  product_cd= c(1, 2, 3, 4, 6),
  description= c("Soft drink", "Frozen snack", "Fruit", "Water", "Fruit 2"),
  unit_price= c(3.0, 5.0, 5.0, 2.0, 4.0), 
  unit_measure= c("each", "each", "kg", "each", "kg")
  )
```

#### 4.9.1 Left join

All the rows from sales and matched rows from products.

```{r left_join}
sales %>% left_join(products, by="product_cd")
```

#### 4.9.2 Right join

All the rows from stores and matched rows from sales.

```{r right_join}
sales %>% right_join(stores, by="store_cd")
```

#### 4.9.3 Inner join

All the rows common to sales and stores.

```{r inner_join}
sales %>% inner_join(stores, by="store_cd")
```

#### 4.9.4 Full join

All the rows from sales and stores.

```{r full_join}
sales %>% full_join(stores)
```

#### 4.9.5 Anti-join

Only rows that are in sales but not in products.

```{r anti_join}
sales %>% anti_join(products)
```

### 4.10 Reshaping tables

This is useful when you do not have a tidy data set. A tidy data set if often required by plotting functions and many data exploration functions. Let us look at this example:

```{r data2}
df_wide <- data.frame(
  project = c("project1", "project2", "project3"),
  Jan= sample(1000:2000, 3),
  Feb= sample(1000:2000, 3),
  Mar= sample(1000:2000, 3)
)
```

This is not a tidy data set (there is more than 1 observation per row). We can convert it to a tidy data set using the pivot function.

```{r wide_long}
df_long <- df_wide %>% 
  pivot_longer(cols= 2:4,
               names_to = "months",
               values_to = "expenses")
```

If we need a smaller table for visualization or for a presentation, we can pivot it back to the original format that is more compact and easier to read for your audience.

```{r long_wide}
df_wide_2 <- df_long %>%
  pivot_wider(names_from = "months",
              values_from = "expenses")
```
