---
title: "Tidyverse tutorial 2 - More advanced operations"
author: "Ariane Ducellier"
date: "10/03/2024"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load R packages.

```{r libraries}
library(httr)
library(jsonlite)
library(mice)
library(rvest)
library(tidyverse)
```

## 1. Dealing with missing data

```{r get_data}
header <- c("age", "workclass", "fnlwgt", "education",
  "education_num", "marital_status", "occupation",
  "relationship", "race", "sex", "capital_gain",
  "capital_loss", "hours_per_week", "native_country", "target")
df <- read_csv("https://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.data",
  col_names=header, trim_ws=TRUE)
df <-df %>%
  mutate(workclass = na_if(workclass, "?"),
         occupation = na_if(occupation, "?"),
         native_country = na_if(native_country, "?"))
```

### 1.1 Filling values with previous value

```{r fill_previous}
df_fill1 <- df %>%
  fill(workclass, occupation, native_country, .direction="down")
```

### 1.2 Filling values with most frequent value

For categorical variables.

```{r fill_frequent}
m_freq_workcls <- names(table(df$workclass))[which.max(table(df$workclass))]
m_freq_occup <- names(table(df$occupation))[which.max(table(df$occupation))]
df_fill2 <- df %>%
  replace_na(list(workclass = m_freq_workcls,
                  occupation = m_freq_occup))
```

### 1.3 Dropping rows with missing values

Dropping rows with at least one missing value.

```{r omit}
df_no_na <- df %>% na.omit()
```

Dropping rows with missing values for specific columns.

```{r drop}
df_native <- df %>%
  drop_na(native_country)
```

### 1.4 Imputing with mice

```{r data}
data("txhousing")
txhousing$date <- date_decimal(txhousing$date, tz="GMT")
txhousing$city <- as.factor(txhousing$city)
```

```{r drop_5nas}
idx <- which(rowSums(is.na(txhousing)) == 5)
txhousing <- txhousing[-idx,]
```

Impute median value for sales, volume and median.

```{r impute_median}
txhousing$sales[is.na(txhousing$sales)] <- median(txhousing$sales, na.rm=TRUE)
txhousing$volume[is.na(txhousing$volume)] <- median(txhousing$volume, na.rm=TRUE)
txhousing$median[is.na(txhousing$median)] <- median(txhousing$median, na.rm=TRUE)
```

Use mice to impute listings and inventory.

```{r impute_mice}
impute <- mice(data.frame(txhousing[,7:8]), seed=123)
impute_data <- complete(impute, 1)
txhousing_clean <- txhousing %>%
  mutate(listings = impute_data[,1],
         inventory = impute_data[,2])
```

### 1.5 Implicit missing values

The price for the 1st quarter of 2021 is missing, but you won't see it but just looking for the rows with NA.

```{r data_missing}
stocks <- tibble(
  year  = c(2020, 2020, 2020, 2020, 2021, 2021, 2021),
  qtr   = c(   1,    2,    3,    4,    2,    3,    4),
  price = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
```

It becomes obvious when you pivot to a wider table, but then you no longer have tidy data.

```{r wider}
stocks %>%
  pivot_wider(
    names_from = qtr, 
    values_from = price
  )
```

the complete function will fill your tidy dataset with the missing rows and NA for the missing value:

```{r complete}
stocks %>% complete(year, qtr)
```

## 2. Getting data from the web

- Go to the [Wiki page](https://en.wikipedia.org/wiki/List_of_countries_by_GDP_(nominal)).
- Right-click and select Inspect.
- Find the piece of code that highlights the table.
- Right-click and select Copy > XPath.

```{r read_page}
page <- "https://en.wikipedia.org/wiki/List_of_countries_by_GDP_(nominal)"
gdp <- rvest::read_html(page)
```

Get the first paragraph.

```{r get_paragraph}
p1 <- gdp %>%
  html_elements("p") %>%
  html_text()
p1[3]
```

Get the table.

```{r get_table}
gdp_df <- gdp %>%
  html_elements(xpath = '//*[@id="mw-content-text"]/div[1]/table[2]') %>%
  html_table() %>%
  .[[1]]
```

## 3. Getting data from an API

The base URL is: https<nolink>://api.fiscaldata.treasury.gov/services/api/fiscal_service

The end point is: /v1/accounting/mts/mts_table_1

Gathering both gives you [data](https://api.fiscaldata.treasury.gov/services/api/fiscal_service/v1/accounting/mts/mts_table_1) in the JSON format.

```{r url}
url <- "https://api.fiscaldata.treasury.gov/services/api/fiscal_service/v1/accounting/mts/mts_table_1"
treasury_api <- GET(url)
```

```{r json_to_df}
result <- content(treasury_api, "text", encoding="UTF-8")
df_json <- fromJSON(result, flatten=TRUE)
df <- as.data.frame(df_json$data)
```

## 4. Miscellaneous functions

To apply the same function to all the columns in the data set:

```{r map}
mtcars %>% 
  select(hp, wt) %>%  map(mean)
```

To combine data sets by rows:

```{r bind_rows}
A <- mtcars[1:3, ]
B <- mtcars[4:6, ]
AB <- A %>% bind_rows(B)
```

To combine data sets by columns:

```{r bind_cols}
A <- mtcars[1:5, 1:3]
B <- mtcars[1:5, 4:6]
 AB <- A %>% bind_cols(B)
```

Another way or creating a new column with a condition. It allows handling multiple cases of logical tests.

```{r case_when}
mtcars %>% 
  mutate(transmission_type = 
            case_when(
              am == 0 ~ "automatic",
              am == 1 ~ "manual"))
```
